---
title: "assignment7_try2"
format: html
---

## henry 
```{r}
pars <- c(rprey=0.95, alpha=0.01, eff=0.6, pmort=0.4, K=2000, h_prey=0.1)
currpop <- c(prey=100, pred=10)
days <- seq(1, 500)

res <- ode(func=lotvmodk_hunt, y=currpop, times=days, parms=pars)
resl <- as.data.frame(res) %>% pivot_longer(-time, names_to="species", values_to="pop")
ggplot(resl, aes(time, pop, col=species)) + geom_line()

# stability testing 
h_prey <- seq(0, 0.5, by = 0.05) # Test 10 different hunting rates from 0 to 50%
n_sims <- 100 
noise_sd <- 0.10 # 10% variation will be used on each parameter
prey_threshold <- 50 # Stability threshold for prey
pred_threshold <- 10 # Stability threshold for pred
base_pars <- c(rprey=0.95, alpha=0.01, eff=0.6, pmort=0.4, K=2000)
days <- seq(1, 365 * 50, by = 10) # Sample every 10 days for 50 years

results <- map_dfr(h_prey, function(h) {
  
  stable <- replicate(n_sims, {
    # Create parameters with uncertainty
    noisy_pars <- rnorm(length(base_pars),
                        mean = base_pars,
                        sd = abs(base_pars * noise_sd))
    noisy_pars <- pmax(noisy_pars, 0)
    
    names(noisy_pars) <- names(base_pars)
    
    pars <- c(noisy_pars, h_prey = h, min_prey = 50)
    
# Output   
out <- ode(func = lotvmodk_hunt, y = c(prey=1, pred=1), times = days, parms = pars)

final_prey <- tail(out[, "prey"], 1)
final_pred <- tail(out[, "pred"], 1)

final_prey > prey_threshold & final_pred > pred_threshold
  })
  
  data.frame(hunt_rate = h, prop_stable = mean(stable))
})


# Plot
ggplot(results, aes(hunt_rate, prop_stable)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 0.80, lty = 2, col = "red") + # If 80% of returns are stable, I'll consider it to be a safe hunting rate
  labs(x = "Hunting Rate",
       y = "Proportion of Stable Runs",
       title = "Predator & Prey Stability Under Hunting Pressure",
       caption = "Dashed line = 80% stability threshold.\nStability defined as prey > 50 and pred > 10 at year 50.") +
  theme_minimal()
```


# Try again 
```{r}
# inital conditions
inital_cond <- c(prey = 100, pred = 20)

# time points
tm <- seq(from = 1, to = 200) # for 200 days 

# set parameters 
pars <- c(rprey = 0.95, alpha = 0.01, eff = 0.6, pmort = 0.5, K = 2000, h_prey = 0.2)

# run the model
res <- ode(func = lotvmodk_hunt, y = inital_cond, times = tm, parms = pars)

```

```{r}
# Look at the results 
head(res)

# Make results tidy data 
res_df <- as.data.frame(res) %>% 
  pivot_longer(-time, 
               names_to = "animal", 
               values_to = "pop")
head(res_df)

# Graph 
ggplot(res_df, aes(time, pop, col = animal)) +
  geom_line()

ggplot(as.data.frame(res), aes(pred, prey, col = time)) + 
  geom_point() +
  labs(y = "Prey", x = "Predators")
```

The plots above shows the prey and predator populations oscillating overtime as parameters (including hunting) are held constant.


## Run Sobol 
To understand how population estimates vary with varying parameters. 

**Generate Parameters**
rprey=0.95, alpha=0.01, eff=0.6, pmort=0.4, K=2000 with a standard deivation of 10%. 
```{r}
np <- 200

K <- rnorm(mean = 2000, sd = 200, n = np)
rprey <- rnorm(mean = 0.95, sd = 0.095, n = np)
alpha <- rnorm(mean = 0.01, sd = 0.001, n = np)
eff <- rnorm(mean = 0.6, sd = 0.06, n = np)
pmort <- rnorm(mean = 0.4, sd = 0.04, n = np)
h_prey <- rnorm(mean = 0.3, sd = 0.03, n = np)

x1 <- cbind.data.frame(rprey = rprey, K = K, alpha = alpha, eff = eff, pmort = pmort, h_prey = h_prey)

# repeat to get our second set of samples
K <- rnorm(mean = 2000, sd = 200, n = np)
rprey <- rnorm(mean = 0.95, sd = 0.095, n = np)
alpha <- rnorm(mean = 0.01, sd = 0.001, n = np)
eff <- rnorm(mean = 0.6, sd = 0.06, n = np)
pmort <- rnorm(mean = 0.4, sd = 0.04, n = np)
h_prey <- rnorm(mean = 0.3, sd = 0.03, n = np)

x2 <- cbind.data.frame(rprey = rprey, K = K, alpha = alpha, eff = eff, pmort = pmort, h_prey = h_prey)

```

**Create Sobol Object**
```{r}
# Create sobol object 
sens_pp <- sobolSalt(model = NULL, x1, x2, nboot = 300)

# Change column names 
colnames(sens_pp$X) <- c("rprey", "K", "alpha", "eff", "pmort", "h_prey")

```

**Create metric function**
```{r}
# Create a function to get max and min of 
# compute_metrics <- function(result) {
#   maxprey <- max(result$prey)
#   maxpred <- max(result$pred)
#   minprey <- min(result$prey)
#   minpred <- min(result$pred)
#   #prey_stable <- as.numeric(min(result$prey) > 0.1 * K) # 1 stable, 0 collapse 
#   return(list(maxprey = maxprey, minprey = minprey, maxpred = maxpred, minpred = minpred)) #prey_stable = prey_stable))
# }


compute_metrics <- function(result) {
 
  meanrprey <- mean(result$prey)
  meanrpred <- mean(result$pred)

  # return results
  return(list(meanrprey = meanrprey, meanrpred = meanrpred))
}
  
```

Create wrapper function

```{r}
wrapper_func <- function(t, pop, rprey, alpha, eff, pmort, K, h_prey, func) {
  parms <- list(rprey = rprey, alpha = alpha, eff = eff, pmort = pmort, K = K, h_prey = h_prey)
  result <- ode(y = pop, times = t, func = func, parms = parms)
  colnames(result) <- c("time", "prey", "pred")
  # get metrics
  metrics <- compute_metrics(as.data.frame(result))
  return(metrics)
}
```

Run Wrapper for all parameters

```{r}
# Set intital pop size and time (200 yrs)
pop <- c(prey = 100, pred = 80)
tm <- seq(from = 1, to = 200)

# Run the model for all parameters 
allresults <- as.data.frame(sens_pp$X) %>% 
    pmap(wrapper_func, pop = pop, t = tm, func = lotvmodk_hunt)
allresults

# Results -> unlisted form
#allres <- allresults %>% map_dfr(`[`, c("maxprey", "minprey", "maxpred", "minpred")) #"prey_stable"))
allres <- allresults %>% map_dfr(`[`, c("meanrprey", "meanrpred")) #"prey_stable"))
allres

# Range of response across parameter uncertainity 
allresl <- allres %>% gather(key = "metric", value = "pop")
allresl


# Create df with parameters and sobol results
allresp_df <- as.data.frame(cbind(sens_pp$X, allres))
allresp_df
```

Graphs

```{r}
ggplot(allresl, aes(metric, pop)) +
  geom_boxplot()

ggplot(allresl, aes(metric, pop, col = metric)) +
  geom_boxplot() +
  facet_wrap(~metric, scales = "free")

ggplot(allresl, aes(pop, col = metric)) +
  stat_ecdf(geom = "line") +
  facet_wrap(~metric, scales = "free")

```

Sobol Indices

```{r}
# for minprey 
#sens_PP_minprey <- sens_pp %>% sensitivity::tell(y = allres$minprey)
sens_PP_minprey <- sens_pp %>% sensitivity::tell(y = allres$meanrprey)


rownames(sens_PP_minprey$S) <- c("rprey", "K", "alpha", "eff", "pmort", "h_prey")
rownames(sens_PP_minprey$T) <- c("rprey", "K", "alpha", "eff", "pmort", "h_prey")

# for maxprey 
#meanrpred
sens_PP_maxprey <- sens_pp %>% sensitivity::tell(y = allres$meanrpred)
# sens_PP_maxprey <- sens_pp %>% sensitivity::tell(y = allres$maxprey)
rownames(sens_PP_maxprey$S) <- c("rprey", "K", "alpha", "eff", "pmort", "h_prey")
rownames(sens_PP_maxprey$T) <- c("rprey", "K", "alpha", "eff", "pmort", "h_prey")

# for minpred
# sens_PP_minpred <- sens_pp %>% sensitivity::tell(y = allres$minpred)
# rownames(sens_PP_minpred$S) <- c("rprey", "K", "alpha", "eff", "pmort", "h_prey")
# rownames(sens_PP_minpred$T) <- c("rprey", "K", "alpha", "eff", "pmort", "h_prey")
# 
# # for maxpred
# sens_PP_maxpred <- sens_pp %>% sensitivity::tell(y = allres$maxpred)
# rownames(sens_PP_maxpred$S) <- c("rprey", "K", "alpha", "eff", "pmort", "h_prey")
# rownames(sens_PP_maxpred$T) <- c("rprey", "K", "alpha", "eff", "pmort", "h_prey")

```

Put Sobol Indices together in one table

```{r}
# S Indices 
sobol_S = as.data.frame(
   rbind(
    sens_PP_maxprey$S$original,
    sens_PP_minprey$S$original,
   #sens_PP_maxpred$S$original,
    #sens_PP_minpred$S$original
  ))
colnames(sobol_S) <- c("rprey", "K", "alpha", "eff", "pmort", "h_prey")
rownames(sobol_S) <- c("maxprey", "minprey", "maxpred", "minpred")
sobol_S

# T Indices 
sobol_T = as.data.frame(
   rbind(
    sens_PP_maxprey$T$original,
    sens_PP_minprey$T$original,
   sens_PP_maxpred$T$original,
    sens_PP_minpred$T$original
  ))
colnames(sobol_T) <- c("rprey", "K", "alpha", "eff", "pmort", "h_prey")
rownames(sobol_T) <- c("maxprey", "minprey", "maxpred", "minpred")
sobol_T
```

The first order sensitivity results show vary different 
The variance associated with variable interactions is mostly due to the growth rate of the prey population (`rprey`) and the interaction coefficient (`alpha`). Hunting rate was not the main cause of any of the output variances. 

```{r}

ggplot(allresp_df, aes(h_prey, minprey, col=pmort )) +
  geom_point() +
  theme_minimal()


```


The rate of maximum change of prey populations are most sensitive to the parameter `K` for the first order sensitivity indices. The rate of maximum change of prey populations are most sensitive to the parameter `rhunt` for the total sensitivity analysis. The rate of hunting is directly reducing the number of prey. The rate of maximum change of prey populations is secondly most sensitive to the `rprey` parameter in the first order indices. The rate of maximum change of prey populations is secondly most sensitive to the `pmort` parameter in the total order indices. This is the mortality rate of the predator population. 

I chose to look at the mean rate of change of prey because these would show the average scenario for the populations of each animal. This is an important metric to give a realistic look at the populations of both animal groups. 




