---
title: "assignment7_try4"
format: html
---

---
title: "Assignment 7 - Two Variables ODE"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

------------------------------------------------------------------------

# Assignment 7

Consider how you might add hunting of prey to the predator prey model that we've been using in class

```{r}
library(sensitivity)
library(tidyverse)
library(deSolve)
library(lhs)
library(purrr)
library(here)
library(ggpubr)
```

## **Part 1**:

Build this model (e.g add hunting to the lotvmodK.R),

Some requirements/hints for your model

You should make sure that you don't hunt more prey than exist.

To ensure that you might also add a minimum prey population input that must be met before hunting is allowed.

Note you can make this as simple or as complex as you would like. You could represent hunting in a way that is similar to "harvesting" in the last assignment - or you could make hunting an exogenous process - so a time based input to your model

```{r}
source(here::here("assignment7", "lotvmodk_hunt.R"))
lotvmodk_hunt
```

## **Part 2**

Explore how different hunting levels and uncertainty/variation in parameters in your model are likely to effect the stability of the populations of both predator and prey. A key challenge is how you might want to define stability? It is up to you but you will need to write a sentence to explain why you chose the measure that you did. It could be something as simple as maintaining a population above some value 50 years into the future. 

[**Stability:**]{.underline}

-   Systems ability to maintain a balanced population of both prey and predator with healthy oscillations between the populations.

-   The prey population is stable if it never drops below 10% of the carrying capacity (k) over 100 years. This definitions allows stability to be relative to the system and not an absolute/arbitrary number.

-   The prey population is stable if it never drops below 50% of the original population size over 100 years. Here, the initial prey population is 100. Therefore, the prey population is considered un-stable when the population numbers decrease below 50.

Use this exploration to recommend a hunting target that will be sustainable (e.g leave you with a stable prey and predator population).

It is up to you how you "explore" hunting - you can simply try different values of the parameters in your hunting model or do it more formally by running your model across a range of values. You could think about parameter interactions

You can assume the following are best guesses of key parameters

rprey=0.95, alpha=0.01, eff=0.6,pmort=0.4, K=2000,

### Implementation of Predation-Prey function (with hunting)

Looking at predator and prey population with 10% predatory and prey hunting rates.

```{r}
# inital conditions
inital_cond <- c(prey = 100, pred = 20)

# time points
tm <- seq(from = 1, to = 200) # for 200 days 

# set parameters 
pars <- c(rprey = 0.95, alpha = 0.01, eff = 0.6, pmort = 0.5, K = 2000, h_prey = 0.2)

# run the model
res <- ode(func = lotvmodk_hunt, y = inital_cond, times = tm, parms = pars)

```

```{r}
# Look at the results 
head(res)

# Make results tidy data 
res_df <- as.data.frame(res) %>% 
  pivot_longer(-time, 
               names_to = "animal", 
               values_to = "pop")
head(res_df)

# Graph 
ggplot(res_df, aes(time, pop, col = animal)) +
  geom_line()

ggplot(as.data.frame(res), aes(pred, prey, col = time)) + 
  geom_point() +
  labs(y = "Prey", x = "Predators")
```

### Run sobol

Generate Parameters

rprey=0.95, alpha=0.01, eff=0.6,pmort=0.4, K=2000 And sd of 10%

```{r}
np <- 200

K <- rnorm(mean = 2000, sd = 200, n = np)
rprey <- rnorm(mean = 0.95, sd = 0.095, n = np)
alpha <- rnorm(mean = 0.01, sd = 0.001, n = np)
eff <- rnorm(mean = 0.6, sd = 0.06, n = np)
pmort <- rnorm(mean = 0.4, sd = 0.04, n = np)
h_prey <- rnorm(mean = 0.3, sd = 0.03, n = np)

x1 <- cbind.data.frame(rprey = rprey, K = K, alpha = alpha, eff = eff, pmort = pmort, h_prey = h_prey)

# repeat to get our second set of samples
K <- rnorm(mean = 2000, sd = 200, n = np)
rprey <- rnorm(mean = 0.95, sd = 0.095, n = np)
alpha <- rnorm(mean = 0.01, sd = 0.001, n = np)
eff <- rnorm(mean = 0.6, sd = 0.06, n = np)
pmort <- rnorm(mean = 0.4, sd = 0.04, n = np)
h_prey <- rnorm(mean = 0.3, sd = 0.03, n = np)

x2 <- cbind.data.frame(rprey = rprey, K = K, alpha = alpha, eff = eff, pmort = pmort, h_prey = h_prey)
```

Create Sobol Object

```{r}
# Create sobol object 
sens_pp <- sobolSalt(model = NULL, x1, x2, nboot = 300)

# Change column names 
colnames(sens_pp$X) <- c("rprey", "K", "alpha", "eff", "pmort", "h_prey")
sens_pp$X
```

Create metric function

```{r}
# Create a function to get mean prey and predators 
compute_metrics <- function(result) {
 
  meanprey <- mean(result$prey)
  meanpred <- mean(result$pred)

  # return results
  return(list(meanprey = meanprey, meanpred = meanpred))
}
```

Create wrapper function

```{r}
wrapper_func <- function(t, pop, rprey, alpha, eff, pmort, K, h_prey, h_pred, func) {
  parms <- list(rprey = rprey, alpha = alpha, eff = eff, pmort = pmort, K = K, h_prey = h_prey, h_pred = h_pred)
  result <- ode(y = pop, times = t, func = func, parms = parms)
  colnames(result) <- c("time", "prey", "pred")
  # get metrics
  metrics <- compute_metrics(as.data.frame(result))
  return(metrics)
}
```

Run Wrapper for all parameters

```{r}
# Set intital pop size and time (200 yrs)
pop <- c(prey = 100, pred = 80)
tm <- seq(from = 1, to = 200)

# Run the model for all parameters 
allresults <- as.data.frame(sens_pp$X) %>% 
    pmap(wrapper_func, pop = pop, t = tm, func = lotvmodk_hunt)

# Results -> unlisted form
allres <- allresults %>% map_dfr(`[`, c("meanprey", "meanpred")) #"prey_stable"))
allres

# Range of response across parameter uncertainity 
allresl <- allres %>% gather(key = "metric", value = "pop")
allresl

# Create df with parameters and sobol results
allresp_df <- as.data.frame(cbind(sens_pp$X, allres))
allresp_df
```

Graphs

```{r}
ggplot(allresl, aes(metric, pop)) +
  geom_boxplot()

ggplot(allresl, aes(metric, pop, col = metric)) +
  geom_boxplot() +
  facet_wrap(~metric, scales = "free")

ggplot(allresl, aes(pop, col = metric)) +
  stat_ecdf(geom = "line") +
  facet_wrap(~metric, scales = "free")

```

Sobol Indices

```{r}
# for minprey 
sens_PP_meanprey <- sens_pp %>% sensitivity::tell(y = allres$meanprey)

rownames(sens_PP_meanprey$S) <- c("rprey", "K", "alpha", "eff", "pmort", "h_prey")
rownames(sens_PP_meanprey$T) <- c("rprey", "K", "alpha", "eff", "pmort", "h_prey")

# for meanpred 
sens_PP_maxprey <- sens_pp %>% sensitivity::tell(y = allres$meanpred)
rownames(sens_PP_maxprey$S) <- c("rprey", "K", "alpha", "eff", "pmort", "h_prey")
rownames(sens_PP_maxprey$T) <- c("rprey", "K", "alpha", "eff", "pmort", "h_prey")


```

But Sobol Indices together in one table

```{r}
# S Indices 
sobol_S = as.data.frame(
   rbind(
    sens_PP_maxprey$S$original,
    sens_PP_minprey$S$original,
  ))
colnames(sobol_S) <- c("rprey", "K", "alpha", "eff", "pmort", "h_prey", "h_pred")
rownames(sobol_S) <- c("maxprey", "minprey", "maxpred", "minpred")
sobol_S

# T Indices 
sobol_T = as.data.frame(
   rbind(
    sens_PP_maxprey$T$original,
    sens_PP_minprey$T$original,
   sens_PP_maxpred$T$original,
    sens_PP_minpred$T$original
  ))
colnames(sobol_T) <- c("rprey", "K", "alpha", "eff", "pmort", "h_prey", "h_pred")
rownames(sobol_T) <- c("maxprey", "minprey", "maxpred", "minpred")
sobol_T
```

According to the Sobol indices, alpha is dominating the variance across all metrics. Whereas, hunting rates have essentially no influence.

Since I am interesting in understanding hunting's influences on predator-prey populations, I will fix the alpha at its mean value (0.01) to remove its dominance.

```{r}
alpha <- rep(0.01, np) # Fixed at the mean 

K <- rnorm(mean = 2000, sd = 200, n = np)
rprey <- rnorm(mean = 0.95, sd = 0.095, n = np)
eff <- rnorm(mean = 0.6, sd = 0.06, n = np)
pmort <- rnorm(mean = 0.4, sd = 0.04, n = np)
h_prey <- rnorm(mean = 0.1, sd = 0.01, n = np)
h_pred <- rnorm(mean = 0.1, sd = 0.01, n = np)

x1 <- cbind.data.frame(rprey = rprey, K = K, alpha = alpha, eff = eff, pmort = pmort, h_prey = h_prey, h_pred = h_pred)

# repeat to get our second set of samples
K <- rnorm(mean = 2000, sd = 200, n = np)
rprey <- rnorm(mean = 0.95, sd = 0.095, n = np)
eff <- rnorm(mean = 0.6, sd = 0.06, n = np)
pmort <- rnorm(mean = 0.4, sd = 0.04, n = np)
h_prey <- rnorm(mean = 0.1, sd = 0.01, n = np)
h_pred <- rnorm(mean = 0.1, sd = 0.01, n = np)

x2 <- cbind.data.frame(rprey = rprey, K = K, alpha = alpha, eff = eff, pmort = pmort, h_prey = h_prey, h_pred = h_pred)

```

```{r}
# Create sobol object 
sens_pp <- sobolSalt(model = NULL, x1, x2, nboot = 300)

# Change column names 
colnames(sens_pp$X) <- c("rprey", "K", "alpha", "eff", "pmort", "h_prey", "h_pred")

# Set intital pop size and time (200 yrs)
pop <- c(prey = 100, pred = 80)
tm <- seq(from = 1, to = 200)

# Run the model for all parameters 
allresults <- as.data.frame(sens_pp$X) %>% 
    pmap(wrapper_func, pop = pop, t = tm, func = pred_prey_func)

# Results -> unlisted form
allres <- allresults %>% map_dfr(`[`, c("maxprey", "minprey", "maxpred", "minpred")) #"prey_stable"))

# Range of response across parameter uncertainity 
allresl <- allres %>% gather(key = "metric", value = "pop")

```

Sobol Indices

```{r}
# for minprey 
sens_PP_minprey <- sens_pp %>% sensitivity::tell(y = allres$minprey)

rownames(sens_PP_minprey$S) <- c("rprey", "K", "alpha", "eff", "pmort", "h_prey", "h_pred")
rownames(sens_PP_minprey$T) <- c("rprey", "K", "alpha", "eff", "pmort", "h_prey", "h_pred")

# for maxprey 
sens_PP_maxprey <- sens_pp %>% sensitivity::tell(y = allres$maxprey)
rownames(sens_PP_maxprey$S) <- c("rprey", "K", "alpha", "eff", "pmort", "h_prey", "h_pred")
rownames(sens_PP_maxprey$T) <- c("rprey", "K", "alpha", "eff", "pmort", "h_prey", "h_pred")

# for minpred
sens_PP_minpred <- sens_pp %>% sensitivity::tell(y = allres$minpred)
rownames(sens_PP_minpred$S) <- c("rprey", "K", "alpha", "eff", "pmort", "h_prey", "h_pred")
rownames(sens_PP_minpred$T) <- c("rprey", "K", "alpha", "eff", "pmort", "h_prey", "h_pred")

# for maxpred
sens_PP_maxpred <- sens_pp %>% sensitivity::tell(y = allres$maxpred)
rownames(sens_PP_maxpred$S) <- c("rprey", "K", "alpha", "eff", "pmort", "h_prey", "h_pred")
rownames(sens_PP_maxpred$T) <- c("rprey", "K", "alpha", "eff", "pmort", "h_prey", "h_pred")

# Put indicies together 
# S indicies
sobol_S2 = as.data.frame(
   rbind(
    sens_PP_maxprey$S$original,
    sens_PP_minprey$S$original,
   sens_PP_maxpred$S$original,
    sens_PP_minpred$S$original
  ))
colnames(sobol_S2) <- c("rprey", "K", "alpha", "eff", "pmort", "h_prey", "h_pred")
rownames(sobol_S2) <- c("maxprey", "minprey", "maxpred", "minpred")
sobol_S2

# T indicies
sobol_T2 = as.data.frame(
   rbind(
    sens_PP_maxprey$T$original,
    sens_PP_minprey$T$original,
   sens_PP_maxpred$T$original,
    sens_PP_minpred$T$original
  ))
colnames(sobol_T2) <- c("rprey", "K", "alpha", "eff", "pmort", "h_prey", "h_pred")
rownames(sobol_T2) <- c("maxprey", "minprey", "maxpred", "minpred")
sobol_T2
```

Now, eff is dominating min_prey values, followed by pmort. Hunting rates on prey still have very limited effect on minimal prey populations. Minimum predator populations are mostly influenced by rprey which is the preys' growth rate.


To get a better understanding of hunting rates, I investigate min prey and min predatory populations in relationship to varying hunting rates.

```{r}
# Set intital pop size and time (200 yrs)
pop <- c(prey = 100, pred = 80)
tm <- seq(from = 1, to = 200)

# Vary hunting rate 
h_prey_test <- seq(0, 0.9, by = 0.01)  # 0% to 90% hunting rate

# Look at hunthing when all other parameters are fixed 
results_sweep <- map_dfr(h_prey_test, function(h) {
  pars <- c(rprey = 0.95, alpha = 0.01, eff = 0.6, pmort = 0.5, # Fixed parameters
            K = 2000, h_prey = h, h_pred = 0.1)
  res <- as.data.frame(ode(y = pop, times = tm, func = pred_prey_func, parms = pars)) # Run ode and pred_prey_func
  colnames(res) <- c("time", "prey", "pred")
  data.frame(h_prey = h, minprey = min(res$prey), minpred = min(res$pred))
})

results_sweep

# Plot
ggplot(results_sweep, aes(x = h_prey)) +
  geom_line(aes(y = minprey), color = "pink") +
  geom_line(aes(y = minpred), color = "blue") +
  #geom_hline(yintercept = 0.1 * 2000, linetype = "dashed") +  # stability threshold
  geom_hline(yintercept = 50, linetype = "dashed") + # stability threshold
  labs(title = "Predator and Prey Populations with Various Hunting rates", 
       x = "Hunting rate (h_prey)", 
       y = "Minimum population") + 
  theme_bw()
```

Around a hunting pressure of \~12%, predator and prey populations begin to degrade. Therefore, a suggested hunting rate would be around 15% to avoid major population decreases.

Submit - the Quato document with your analysis AND a pdf/html version, any R files This should include

a)  your hunting model

b)  your exploration (e.g how you tested different hunting levels and how you defined a stability metric

c)  provides you estimated sustainable hunting level and brief explanation of why you chose the metric you did

# Grading

Your grade will be based on the following criteria (note for alternative assignment breakdown will be adapted to fit model)

| Criteria | Full Marks | No Marks | Points |
|-----------------|--------------------|-----------------|-----------------|
| **Adding hunting to pred-prey differential equation** <br> - Correct implementation (including adding parameters and modifying the equation) <br> - Includes a parameter to restrict hunting if prey falls below some value | 3 pts | 0 pts | 3 pts |
| **Developed a metric to quantify stability** <br> - The metric captures some aspect of stability <br> - The metric is implemented as a function that can be applied to a time series (or multiple time series) | 3 pts | 0 pts | 3 pts |
| **Exploration of stability** <br> - Compared the stability metric across different values of parameters used to define hunting <br> - Comparison is explicit in the Rmarkdown (graph or table) showing how hunting levels influence stability | 4 pts | 0 pts | 4 pts |
| **Using results to make conclusions** <br> - 1–2 sentences that justify a decision or policy about the level of hunting that clearly arise from the exploration of stability | 3 pts | 0 pts | 3 pts |
| **Style** <br> - Graphs have clear labels and legends (if needed) and clearly communicate the intended message <br> - Rmarkdown includes text explaining each step leading to the final conclusion about hunting levels | 2 pts | 0 pts | 2 pts |
|  |  | **Total Points** | **15 pts** |
