---
title: 'Assignment: Using Sobol with an ODE'
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Consider the following model of forest growth (where forest size in measured in units of carbon (C))

-   $dC/dt  = r*C$ for forests where C is below a threshold canopy closure

-   $dC/dt = g*(1- C/K)$ for forests where carbon is at or above the threshold canopy closure

-   and $K$ is a carrying capacity in units of carbon

The size of the forest ($C$), Canopy closure threshold and carrying capacity are all in units of carbon


You could think of the canopy closure threshold as the size of the forest at which growth rates change from exponential to linear You can think of $r$, as early exponential growth rate and $g$ as the linear growth rate once canopy closure has been reached

## [Your task]{style="color:orange"} {.scrollable}


5.  Graph the results of the sensitivity analysis as a box plot of maximum forest size and record the two Sobol indices (**S** and **T**).

6.  In 2-3 sentences, discuss what the results of your simulation might mean. (For example think about how what parameters climate change might influence).

Submit rendered Quarto pdf with model implementation, graphs and sensitivity analysis and R file with your model

You can work in groups or individually

## Extra Credit

Compute Sobol indices for a second metric: forest size after a 100 years

## Grading Rubric

```{r}
library(tidyverse)
library(deSolve)
library(here)
library(sensitivity)

# Set the seed 
set.seed(230)
```

1.  Implement this model in R (as a differential equation)

```{r}
# Source function 
source(here::here(("assignment6/forest_growth_func.R")))
```

2.  Run the model for 300 years (using the ODE solver) starting with an initial forest size of 10 kg/C, and using the following parameters:

-   canopy closure threshold of 50 kgC
-   $K$ = 250 kg C (carrying capacity)
-   $r$= 0.01 (exponential growth rate before before canopy closure)
-   $g$ = 2 kg/year (linear growth rate after canopy closure)

3.  Graph the results. Here you are graphing the trajectory with the parameters as given (e.g no uncertainty)

```{r}
# Define Variables 
tm <- seq(from = 1, to = 300)
Cinitial <- 10
gps <- list(r = 0.01, g = 2, K = 250)

# Run ode 
res <- ode(Cinitial, tm, forest_growth_func, gps)

# Change column names 
colnames(res) <- c("time", "carbon")

# Graph 
ggplot(as.data.frame(res), aes(time, carbon)) +
  geom_point()

```

4.  Run a sobol global (vary all parameters at the same time) sensitivity analysis that explores how the estimated **maximum forest size** (e.g maximum of $C$ 300 years, varies with these parameters

-   pre canopy closure growth rate ($r$)
-   post-canopy closure growth rate ($g$)
-   canopy closure threshold and carrying capacity($K$)

Assume that parameters are all normally distributed with means as given above and standard deviation of 10% of mean value



Generate the Sample parameters (r, g, K) with means from above and sd of 10% of the mean values 
```{r}
# Generate sample paramters 

np <- 300
r <- rnorm(mean = 0.01, sd = 0.001, n = np)
g <- rnorm(mean = 2,    sd = 0.2,   n = np)
K <- rnorm(mean = 250,  sd = 25,    n = np)

x1 <- cbind.data.frame(r = r, g = g, K = K)

# Generate sample paramters... again! 
r <- rnorm(mean = 0.01, sd = 0.001, n = np)
g <- rnorm(mean = 2,    sd = 0.2,   n = np)
K <- rnorm(mean = 250,  sd = 25,    n = np)

x2 <- cbind.data.frame(r = r, g = g, K = K)
```

Create the Sobol Object 
```{r}
# Create the sobol object
sens_forest <- sobolSalt(model = NULL, 
          x1, x2, 
          nboot = 300)

# Change the Column names 
colnames(sens_forest$X) <- c("r", "g", "K")

# Look at it 
head(sens_forest$X)
```

Create a Function that shows important metrics 
```{r}
# The metrics: the max forest size (Carbon units)
metricfunc <- function(result) {
  maxC <- max(result$Carbon)
  return(list(maxC = maxC))
}

```

Build a Wrapper Function: 
runs ode and extracts the results for all the parameters at once 
```{r}
wrapper_func <- function(r, g, K, Cinitial, simtimes, odefunc, metricfunc) {
  parms <- list(r = r, g = g, K = K) # List of parameters 
  results <- ode(y = Cinitial,  # Run the ODE function
                   times = simtimes, 
                   func = odefunc, 
                   parms = parms)
  results <- as.data.frame(results) # ODE results as a df 
  colnames(results) <- c("times", "Carbon") 
  # Get the metrics: Max forest size 
  metrics <- metricfunc(results)
  #metrics <- metricfunc(as.data.frame(results))
  return(metrics)  
}

```

Use the wrapper function to explores how the estimated maximum forest size changes as the parameters vary 
```{r}
Cinitial <- 10

# Use pmap to run the wrapper function over the sobol `sens_forest` object 
allresults <- as.data.frame(sens_forest$X) %>% pmap(wrapper_func, 
                                      Cinitial = Cinitial, 
                                      simtimes = tm, 
                                      odefunc = forest_growth_func, 
                                      metricfunc = metricfunc)

#head(allresults) # Look at pmap results 

# Extract the results from pmap into a data frame
allres <- allresults %>% map_dfr(`[`, c("maxC"))

head(allres) # Look at dataframe 
```


5. Graph the results of the sensitivity analysis as a box plot of maximum forest size and record the two Sobol indices (S and T).

Graph with a boxplot 
```{r}
ggplot(allres) + 
  geom_boxplot(aes(x = maxC)) + 
  labs(title = "Maximun Forest Size Boxplot", 
       subtitle = "Understanding forest growth rates.",
       x = "Size of the Forest in Units of Carbon") + 
  theme_gray() +
  theme(axis.ticks.y = element_blank(), 
        axis.text.y = element_blank())
```

Compute the Sobol Indicies 
```{r}
# tell our original Sobol object about the results.
sens_forest_sobol_ind <- sensitivity::tell(sens_forest, allres$maxC)

# Change rownames 
rownames(sens_forest_sobol_ind$S) <- c("r", "g", "K") 
rownames(sens_forest_sobol_ind$T) <- c("r", "g", "K") 

# Look at Sobol Indicies 
sens_forest_sobol_ind$S
sens_forest_sobol_ind$T

```

6. In 2-3 sentences, discuss what the results of your simulation might mean. (For example think about how what parameters climate change might influence).

**Results**

The main effect / first order sensitivity (`sens_forest_sobol_ind$S`) results show all the parameters are impacting the output variance. The carrying capacity (K) is impacting the output variance the most, followed closely by the exponential growth rate (r). 

The total effect (`sens_forest_sobol_ind$T`) results show all parameters impact the variance due to parameter interactions, especially carrying capacity (K) and exponential growth rate (r). 

CHAT GPT 
K and r dominate maximum forest size (highest first and total-order indices), while g (post-canopy growth rate) matters least. This suggests that a forest's ultimate carbon sequestration potential is most sensitive to its carrying capacity and early exponential growth rate. Climate change could impact both — drought and heat stress may reduce K (lower water/nutrient availability capping forest size), while warming-extended growing seasons or CO₂ fertilization could increase r, partially offsetting losses, making these two parameters key targets for climate impact assessments.


EXTRA CREDIT: Compute Sobol indices for a second metric: forest size after a 100 years
```{r}


```

